---
title: "m200_hw5_ahowe"
author: "Andrew Howe"
date: "May 6, 2015"
output: html_document
---

```{r,echo=FALSE,results='hide',message=FALSE}
# Load packages
chooseCRANmirror(ind=88) # go Bruin
packagesToLoad = c("aplpack", "vioplot", "ggplot2", "shiny", "dplyr", "tidyr", "minerva", "lattice", "sqldf", "reshape2", "xtable" )
pkgsIn <- function(x) { require(x,character.only = TRUE) || { install.packages(x,); library(x, character.only = TRUE) } } 
lapply(packagesToLoad, pkgsIn)
# yay! global variables!
deals = 1000
```


## Extra Credit / Set up the data

```{r,echo=FALSE,results='asis'}
# Build data
filename = 'Hospitals.csv'
dir = '/Users/andrewhowe/toSort/m200-stats/homeworks/PSM200Problem_Set_5/'
hospData <- read.csv( file = file.path( dir, filename ), head=TRUE, sep="," )
#
# this makes more sense in SQL
#
# SELECT 
#   SUBSTR(type, 1, 4)  AS status, 
#   location            AS location, 
#   SUM(Counts)         AS total 
# FROM 
#   hospData 
# WHERE 
#   SUBSTR(type, 1, 4) = SUBSTR(type, 1, 4) 
# GROUP BY 
#   location, 
#   status
#
# make a long form summary
hospitalObservations <- sqldf('SELECT SUBSTR(type, 1, 4) AS status, location AS location, SUM(Counts) AS total FROM hospData WHERE SUBSTR(type, 1, 4) = SUBSTR(type, 1, 4) GROUP BY location, status') 
# reform the table
hospitalSummary <- acast( data = hospitalObservations, formula = location~status, value.var = "total")
# fix the row names
#hospitalTable <- hospitalSummary[,-1]
#rownames(hospitalTable) <- hospitalSummary[,1]
print(xtable( hospitalSummary ), type='html', width=10)
```

# Data Analysis

## R's built in fisher.test()

```{r,echo=TRUE,eval=FALSE}
fisher.test( hospitalSummary )
```

This fails, with a complaint about running out of workspace : "Error in fisher.test(hospitalSummary) : FEXACT error 40. Out of workspace."

This probably occurred because the factorials are enormous. Now, we'll simulate the result with Monte-Carlo.

```{r,echo=FALSE}
fisher.test( hospitalSummary, simulate.p.value = TRUE)
```

The results of the Fisher's test suggest a significant association between outcome and hospital location.

##Bootstraping Chi-squared and Kullback-Liebler Divergence

```{r,echo=FALSE}

observed = hospitalSummary  #rbind(c(65,35),c(50,50))
expected = rowSums(observed) %o% colSums(observed) / sum(observed)
KL = sum(expected*log(expected/observed))
X2 = sum(((expected-observed)^2)/expected)

#deals = 100
KL.boot = rep( NA, deals )
X2.boot = rep( NA, deals )

# this code is hella confusing.
for (i in 1:deals) {
  dataDims = dim(observed)
  bootSample = array( data = NA, dim = dataDims)
  # we want to resample the (rownames) with replacement using the 
  # probabilities of observations of each (rowname).
  # To do this, we'll just convert everything to index numbers and
  # get a big list of numbers. Then we'll cut up the list into snippets
  # that match the number of obs. per COLUMN and count the ROWS in each
  # by numerical address. Confused yet? I am.
  #
  # get the long line of scrambled values
  bootData = sample( x = 1:dataDims[1], size = sum(observed), replace=TRUE, prob=rowSums(observed)/sum(observed) )
  # build index cuts for long line of scrambled values
  idxLimitsLong = array( data=NA, dim=c(dataDims[2],2) )
  numObsRows = colSums(observed)
  idxLimitsLong[1,1] = 1
  idxLimitsLong[1,2] = numObsRows[1]
  for ( k in 2: dataDims[2] ) {
    idxLimitsLong[k,1] = 1 + idxLimitsLong[k-1,2]
    idxLimitsLong[k,2] = idxLimitsLong[k,1] + numObsRows[k] - 1
  }
  # cut the big line into column sizes
  for ( colIdx in 1:dataDims[2] ) {
    snip = bootData[idxLimitsLong[colIdx,1]:idxLimitsLong[colIdx,2]]
    for ( rowIdx in 1:dataDims[1] ) {
      bootSample[rowIdx,colIdx] = sum(snip == rowIdx)
    }
  }
  expected.boot = rowSums(bootSample) %o% colSums(bootSample) / sum(bootSample)
  KL.boot[i] = sum(expected.boot*log(expected.boot/bootSample))
  X2.boot[i] = sum(((expected.boot - bootSample)^2)/expected.boot)
}

print(X2)
x2pval=sum(X2.boot>X2)/deals

print(KL)
klpval=sum(KL.boot>KL)/deals
```

The results of the bootstrap indicate that a Chi Square value of `r round(X2, digits = 4)` is highly unlikely with a p-value of `r round(x2pval, digits = 4 )` (zero means the bootstrap cannot accurately estimate how unlikely this observation is.)

The results of the bootstrap indicate that a KL value of `r round(KL, digits = 4 )` is highly unlikely with a p-value of `r round(klpval, digits = 4 )` (zero means the bootstrap cannot accurately estimate how unlikely this observation is.)

##Relative Risk of Death Matrix

```{r,echo=FALSE}
# wooooooooooo
observed = hospitalSummary  #rbind(c(65,35),c(50,50))
pctFatalIntv = observed[,"MORT"]/rowSums(observed)
rrMat <- t(t(pctFatalIntv)) %*% t(1/pctFatalIntv)
```

###Fraction of deaths after interventions for life-threatening conditions

```{r,echo=FALSE,results='asis'}
print( pctFatalIntv )
```

###Relative Risk of Death Matrix

The way to read this is "the relative risk of death in the hospital <row name> is <table value> compared to <column name> (over <column name>).

```{r,echo=FALSE,results='asis'}
print( xtable(rrMat), type='html', width=10 )
```

###Confidence Intervals



###Null Hypothesis Testing P-Values for Risk Ratio 

```{r,echo=FALSE,results='asis'}

observed = hospitalSummary  

#deals = 1000
dataDims = dim(observed)
rrMatBooted <- array( data = rep(NA, dataDims[1]*dataDims[1]*deals), dim = c( dataDims[1], dataDims[1], deals ) )

# this code is hella confusing.
for (i in 1:deals) {
  dataDims = dim(observed)
  bootSample = array( data = NA, dim = dataDims)
  # we want to resample the (rownames) with replacement using the 
  # probabilities of observations of each (rowname).
  # To do this, we'll just convert everything to index numbers and
  # get a big list of numbers. Then we'll cut up the list into snippets
  # that match the number of obs. per COLUMN and count the ROWS in each
  # by numerical address. Confused yet? I am.
  #
  # get the long line of scrambled values
  bootData = sample( x = 1:dataDims[1], size = sum(observed), replace=TRUE, prob=rowSums(observed)/sum(observed) )
  # build index cuts for long line of scrambled values
  idxLimitsLong = array( data=NA, dim=c(dataDims[2],2) )
  numObsRows = colSums(observed)
  idxLimitsLong[1,1] = 1
  idxLimitsLong[1,2] = numObsRows[1]
  for ( k in 2: dataDims[2] ) {
    idxLimitsLong[k,1] = 1 + idxLimitsLong[k-1,2]
    idxLimitsLong[k,2] = idxLimitsLong[k,1] + numObsRows[k] - 1
  }
  # cut the big line into column sizes
  for ( colIdx in 1:dataDims[2] ) {
    snip = bootData[idxLimitsLong[colIdx,1]:idxLimitsLong[colIdx,2]]
    for ( rowIdx in 1:dataDims[1] ) {
      bootSample[rowIdx,colIdx] = sum(snip == rowIdx)
    }
  }
  # perform mathematical transformations
  colnames(bootSample)<-colnames(observed)
  pctFatalIntvBoot = bootSample[,"MORT"]/rowSums(bootSample)
  rrMatBooted[,,i] <- t(t(pctFatalIntvBoot)) %*% t(1/pctFatalIntvBoot)
}

# now plot these things
cn <- colnames(observed)
rn <- rownames(observed)
par(mfrow=c(dataDims[1],dataDims[1]))
#plot.new()
for ( cidx in 1:dataDims[1] ) {
  for ( ridx in 1:dataDims[1] ) {
    if ( cidx == ridx ) {
      #plot(c(1,2),c(2,1), main = rn[ridx] , xlab = '', ylab = '' )
      #text(rn[ridx])
    } else {
      #hist(rrMatBooted[ridx,cidx,], main='', ylab='', xlab='')
      #abline(v = rrMat[ridx,cidx], col='red' )
    }
  }
}
par(mfrow=c(1,1))
```

Above, we have plots of all the histograms for the bootstrapped risk ratios (big-box).

```{r,echo=FALSE,results='asis'}
# Calculate the p-values, both 1 and 2 sided.
pval1side = array( data = rep(NA, dataDims[1]*dataDims[1]), dim = c( dataDims[1], dataDims[1] ) )
colnames(pval1side) <- rownames(observed)
rownames(pval1side) <- rownames(observed)
pval2side = array( data = rep(NA, dataDims[1]*dataDims[1]), dim = c( dataDims[1], dataDims[1] ) )
colnames(pval2side) <- rownames(observed)
rownames(pval2side) <- rownames(observed)
for ( cidx in 1:dataDims[1] ) {
  for ( ridx in 1:dataDims[1] ) {
    if ( cidx == ridx ) {
      pval1side[ridx,cidx] <- NA
    } else {
      if ( rrMat[ridx,cidx] > 1) {
        pval1side[ridx,cidx] <- sum(rrMatBooted[ridx,cidx,]>rrMat[ridx,cidx])/length(rrMatBooted[ridx,cidx,])
        pval2side[ridx,cidx] <- (sum(rrMatBooted[ridx,cidx,]>rrMat[ridx,cidx])+sum(rrMatBooted[ridx,cidx,]<(1/rrMat[ridx,cidx])))/length(rrMatBooted[ridx,cidx,])
      } else {
        pval1side[ridx,cidx] <- sum(rrMatBooted[ridx,cidx,]<rrMat[ridx,cidx])/length(rrMatBooted[ridx,cidx,])
        pval2side[ridx,cidx] <- (sum(rrMatBooted[ridx,cidx,]<rrMat[ridx,cidx])+sum(rrMatBooted[ridx,cidx,]>(1/rrMat[ridx,cidx])))/length(rrMatBooted[ridx,cidx,])
      }
    }
  }
}
print(xtable( pval2side ), type='html', width=10)
```





#References

Healthcare.gov

Made with R 
```{r, echo=FALSE}
version
```